Foundations of ecological and evolutionary change

### 2025-06-18 foundations_of_ecoevo_arxiv.tex
Edits for resubmission of revision to ecology and evolution

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\usepackage{lineno}
\linenumbers
\modulolinenumbers[2]
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Foundations of ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\setstretch{2}
\section{Abstract}\label{abstract}

Biological evolution is realised through the same mechanisms of birth
and death that underlie change in population density. The deep
interdependence between ecology and evolution is well-established, but
much theory in each discipline has been developed in isolation. While
recent work has accomplished eco-evolutionary synthesis, a gap remains
between the logical foundations of ecology and evolution. We bridge this
gap with a new equation that defines a summed value for a characteristic
across individuals in a population, from which the fundamental equations
of population ecology and evolutionary biology (the Price equation) are
derived. We thereby unify the fundamental equations of population
ecology and biological evolution under a general framework. Our
unification further demonstrates the equivalence between mean population
growth rate and evolutionary fitness and links this change to ecosystem
function. Finally, we outline how our proposed framework can be used to
unify social evolution and density-dependent population growth.

@@

Theoretical unification of phenomena is a powerful tool for scientific
advancement. Such unification has been a major goal in scientific
research throughout history (\citeproc{ref-Kitcher1993}{Kitcher 1993}),
and its value is perhaps most evident in reconciling unconnected models
and revealing new and unexpected empirical predictions. In evolutionary
biology, the Price equation (Box 1) provides a unifying framework for
evolutionary theory by exhaustively and exactly describing evolutionary
change for any closed evolving population
(\citeproc{ref-Price1970}{Price 1970}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}). The Price
equation is therefore fundamental, in the sense that it binds together
all of evolutionary theory by formally defining what evolutionary change
is and is not (\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Rice2004}{Rice 2004}; \citeproc{ref-Gardner2008}{Gardner
2008}; \citeproc{ref-Frank2015a}{Frank 2017};
\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Luque2021}{Luque and
Baravalle 2021}). Using this formal definition, the scope of, and
relationships among, sub-disciplines within evolutionary theory can be
clarified. For example, fundamental equations of both population and
quantitative genetics can be derived from the Price equation
(\citeproc{ref-Queller2017}{Queller 2017}). This provides conceptual
clarity by demonstrating the logical consistency of different
theoretical frameworks within evolutionary biology. Our aim here is to
propose an equation that extends this conceptual clarity to include
population change and thereby provide a formal and exact definition of
joint ecological and evolutionary change.

In biological populations, ecological change is caused by the same
processes of individual birth and death that cause changes in allele
frequencies and phenotypes (\citeproc{ref-Turchin2001}{Turchin 2001};
\citeproc{ref-Connor2004}{Connor and Hartl 2004};
\citeproc{ref-Barfield2011}{Barfield et al. 2011}). As with evolution, a
fundamental equation can exhaustively and exactly define population
change. Unlike the Price equation, this fundamental equation is perhaps
self-evident; population change is simply the addition of births and
subtraction of deaths from current population size (\(N_{t}\)) to
recover the new population size (\(N_{t+1}\); Box 2). By definition,
this relationship of \(N_{t+1} = N_{t} + Births - Deaths\) applies to
any closed population. Turchin (\citeproc{ref-Turchin2001}{2001}) argues
that general principles are needed to establish a logical foundation for
population ecology, and this simple birth and death model and the
consequences that logically follow from it (e.g., exponential population
growth) is fundamental to population ecology. Any unifying definition of
joint ecological and evolutionary change must be able to derive both the
Price equation and this birth and death model.

The interdependence of ecological and evolutionary processes has long
been recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
\citeproc{ref-Fisher1958}{Fisher 1958};
\citeproc{ref-Pelletier2009}{Pelletier et al. 2009}), but the rise of
eco-evolutionary models, which incorporate both, is relatively recent
following a widespread recognition that ecology and evolution can happen
on similar timescales (\citeproc{ref-Govaert2019}{Govaert et al. 2019};
\citeproc{ref-Yamamichi2023}{Yamamichi et al. 2023}). Currently, a
universally recognised formal definition of eco-evolutionary change is
lacking, with some theoreticians broadly interpreting ``eco-evolutionary
dynamics'' to allow for a separation of ecological and evolutionary
timescales (\citeproc{ref-Lion2023}{Lion et al. 2023}) and others
advocating for a more narrow interpretation in which no such separation
is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
al. (\citeproc{ref-Bassar2021}{2021}) identify two types of
eco-evolutionary models that follow from these interpretations. The
first type uses separate equations to model population change versus
evolutionary change, thereby allowing for any number of ecological,
evolutionary, or environmental feedbacks (e.g.,
\citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
models population demographics as functions of quantitative traits, with
ecological and evolutionary change following from demographic processes
and trait distributions operating on the same timescale (e.g.,
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Simmonds2020}{Simmonds et al. 2020};
\citeproc{ref-Jaggi2024}{Jaggi et al. 2024}). Both model types can be
very general, but like all predictive models, they rely on simplifying
assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions are
often grounded in the Price equation to demonstrate accuracy and logical
consistency when modelling evolutionary change (e.g.,
\citeproc{ref-Coulson2008}{Coulson and Tuljapurkar 2008};
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Rees2016}{Rees and Ellner 2016};
\citeproc{ref-Lion2018}{Lion 2018}). For example, Barfield et al.
(\citeproc{ref-Barfield2011}{2011}) link their model back to Price
(\citeproc{ref-Price1970}{1970}), which they consider to be a
``universal law of evolution'', to place their conclusions concerning
stage structured evolution in the broader context of evolutionary
theory. The role of fundamental equations is therefore important for
unifying theory (\citeproc{ref-Luque2021}{Luque and Baravalle 2021}),
and we believe that a fundamental equation of eco-evolutionary change is
needed.

We present an equation from which the fundamental equations of ecology
and evolutionary biology can be derived. Derivation follows by adding
assumptions that are specific to population ecology or evolution in the
same way that key equations of population genetics (e.g., average
excess) or quantitative genetics (e.g., Breeder's equation) can be
derived from the Price equation (\citeproc{ref-Queller2017}{Queller
2017}). We propose our equation as a formal definition of
eco-evolutionary change.

@@

\begin{quote}
\textbf{Box 1:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\citeproc{ref-Price1970}{Price 1970},
\citeproc{ref-Price1972}{1972}), it postulates some basic properties
that all evolutionary systems must satisfy: change over time, ancestor
and descendant relations, and a character or phenotype
(\citeproc{ref-Rice2004}{Rice 2004}). Using simple algebraic language,
the Price equation represents evolutionary change with the predominant
notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's fitness, and \(\bar{w}\) average population fitness. On the
right-hand side of the equation, the first term is the covariance
between a character value \(z\) and fitness \(w\), which reflects
\(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Okasha2006}{Okasha 2006}; \citeproc{ref-Frank2012}{Frank
2012}). A more specific version of the covariance term was already known
within the quantitative and population genetics tradition
(\citeproc{ref-Robertson1966}{Robertson 1966}), usually representing the
action of natural selection. The Price equation adds an expectation term
and abstracts away from any specific mechanisms of replication or
reproduction, or mechanisms of inheritance. Its definitional nature and
lack of substantive biological assumptions has been portrayed both as a
strength (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Frank2012}{Frank 2012}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Baravalle2022}{Baravalle and Luque 2022}), and its
greatest weakness. The abstract nature of the Price equation places it
at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fisher's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
\citeproc{ref-Lehtonen2018}{2018}; \citeproc{ref-Queller2017}{Queller
2017}). This abstractness is also key to developing a more general view
of evolution (\citeproc{ref-Rice2020}{Rice 2020};
\citeproc{ref-Luque2021}{Luque and Baravalle 2021};
\citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast, some
researchers consider the Price equation just a triviality (even
tautological), and useless without further modelling assumptions
(\citeproc{ref-VanVeelen2005}{van Veelen 2005};
\citeproc{ref-VanVeelen2012}{van Veelen et al. 2012}). The debate
remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
\citeproc{ref-Baravalle2024}{Baravalle et al. 2025}).
\end{quote}

@@

\begin{quote}
\textbf{Box 2:} The number of individuals in any closed population
(\(N\)) at any given time (\(t + 1\)) is determined by the existing
number of individuals (\(N_{t}\)), plus the number of births
(\(Births\)) minus the number of deaths (\(Deaths\)),
\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
for any closed population. Despite its simplicity, it is a general
equation for defining population change and a starting point for
understanding population ecology. Turchin
(\citeproc{ref-Turchin2001}{2001}) notes that a consequence of this
fundamental equation is the tendency for populations to grow
exponentially (technically geometrically in the above case where time is
discrete). This inherent underlying tendency towards exponential growth
persists even as the complexities of real populations, such as
structure, stochasticity, or density-dependent effects are added to
population models (\citeproc{ref-Turchin2001}{Turchin 2001}). Given the
assumption that all individuals in the population are identical, a per
capita rate of birth, \(Births = bN_{t}\) and death, \(Deaths = dN_{t}\)
can be defined. Rearranging and defining \(\lambda = 1 + b - d\) gives,
\(N_{t+1} = N_{t}\lambda\). Here \(\lambda\) is the finite rate of
increase (\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that
because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
size of any closed population equals its existing size times its finite
rate of increase.
\end{quote}

@@

The foundation of eco-evolutionary change is an interaction between the
demographic processes of birth and death
(\(\beta_{i} - \delta_{i} + 1\)) and some characteristic of an
individual (\(z_{i} + \Delta z_{i}\)). The value of \(\Omega\) is a
summation of individual characteristics, which takes the same units as
\(z\). From eqn 1, we can derive the most fundamental equations of
population ecology (Box 2) and evolutionary biology (Box 1) through an
appropriate interpretation of \(z\). Under some limited interpretations
of \(z\), we can also interpret \(\Omega\) as a metric of ecosystem
function.

@@

To recover the general equation for population ecology (Box 2), we
define \(z_{i}\) as the identity of \(i\) belonging to the population.
In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is a
member of the population. In this case, \(z\) is count, which takes the
unit 1 {[}note that `individual' is a label, not a unit; Newell and
Tiesinga (\citeproc{ref-NewellEtAl2019}{2019}){]}. Further, we assume
that individuals do not change species by setting \(\Delta z_{i} = 0\).
In this case,

@@

We can now interpret \(\Omega\) as the population size at \(t+1\),
\(N_{t+1}\),

@@

We therefore recover the general equation for population ecology (eqn 2)
and the fundamental property of exponential growth in populations
(\citeproc{ref-Turchin2001}{Turchin 2001}) (eqn 3).

@@

In this definition, the longevity of the individual matters. An
individual that survives from \(t\) to \(t + 1\) has a higher fitness
than one that dies, even if both have the same reproductive output. With
this definition of fitness (eqn 4), we substitute,

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\usepackage{lineno}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Foundations of ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\setstretch{1}
\section{Abstract}\label{abstract}

Biological evolution is realised through the same mechanisms of birth
and death that underlie change in population density. The deep
interdependence between ecology and evolution is well-established, and
recent models focus on integrating eco-evolutionary dynamics to
demonstrate how ecological and evolutionary processes interact and feed
back upon each other. Nevertheless, a gap remains between the logical
foundations of ecology and evolution. Both population ecology and
evolution have fundamental equations that define how the size of a
population (ecology) and the average characteristic within a population
(evolution) change, respectively, over time. These fundamental equations
are a complete and exact description of change for any closed
population, but how they are formally linked remains unclear. We link
the fundamental equations of population ecology and evolution with an
equation that sums how individual characteristics interact with
individual fitness in a population. From this equation, we derive the
fundamental equations of population ecology and evolutionary biology
(the Price equation). We thereby identify an overlooked bridge between
ecology and biological evolution. Our unification formally recovers the
equivalence between mean population growth rate and evolutionary fitness
and links this change to ecosystem function. We outline how our
framework can be used to further develop eco-evolutionary theory.

@@

Theoretical unification is a powerful tool for scientific advancement.
Such unification has been a major goal in scientific research throughout
history (\citeproc{ref-Smocovitis1992}{Smocovitis 1992};
\citeproc{ref-Kitcher1993}{Kitcher 1993}), and its value is perhaps most
evident in reconciling unconnected models and revealing new and
unexpected empirical predictions. In evolutionary biology, the Price
equation (Box 1) provides a unifying framework for evolutionary theory
by exhaustively and exactly describing evolutionary change for any
closed evolving population (\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Luque2016}{Luque 2017};
\citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}). The Price equation
is therefore fundamental, in the sense that it binds together all of
evolutionary theory by formally defining what evolutionary change is and
is not (\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Rice2004}{Rice 2004}; \citeproc{ref-Gardner2008}{Gardner
2008}; \citeproc{ref-Frank2015a}{Frank 2017};
\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Luque2021}{Luque and
Baravalle 2021}). Using this formal definition, the scope of, and
relationships among, sub-disciplines within evolutionary theory can be
clarified. For example, fundamental equations of both population and
quantitative genetics can be derived from the Price equation
(\citeproc{ref-Queller2017}{Queller 2017}). This provides conceptual
clarity by demonstrating the logical consistency of different
theoretical frameworks within evolutionary biology. Our aim here is to
propose an equation that extends this conceptual clarity to include
population size change and thereby provide a formal and exact definition
for joint ecological and evolutionary change.

In biological populations, ecological change is caused by the same
processes of individual birth and death that cause evolutionary changes
in allele frequencies and phenotypes (\citeproc{ref-Turchin2001}{Turchin
2001}; \citeproc{ref-Connor2004}{Connor and Hartl 2004};
\citeproc{ref-Barfield2011}{Barfield et al. 2011}). As with evolution, a
fundamental equation can exhaustively and exactly define population
change. Unlike the Price equation, this fundamental equation is perhaps
self-evident. Population change is simply the addition of individuals
minus the removal of individuals from current population size
(\(N_{t}\)), which recovers the new population size (\(N_{t+1}\); Box
2). By definition, this relationship of
\(N_{t+1} = N_{t} + Births - Deaths\) applies to any closed population.
Turchin (\citeproc{ref-Turchin2001}{2001}) argues that general
principles are needed to establish a logical foundation for population
ecology, and this simple birth and death model and the consequences that
logically follow from it (e.g., exponential population growth) are
fundamental to population ecology. Any unifying definition of joint
ecological and evolutionary change must be able to derive both the Price
equation and this birth and death model.

The union of ecological and evolutionary processes has long been
recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
\citeproc{ref-Fisher1958}{Fisher 1958};
\citeproc{ref-Pelletier2009}{Pelletier et al. 2009}), but the rise of
eco-evolutionary models, which incorporate both, is relatively recent
following a widespread recognition that ecology and evolution can happen
on similar timescales (\citeproc{ref-Govaert2019}{Govaert et al. 2019};
\citeproc{ref-Yamamichi2023}{Yamamichi et al. 2023}). Currently, a
universally recognised formal definition of eco-evolutionary change is
lacking, with some theoreticians broadly interpreting ``eco-evolutionary
dynamics'' to allow for a separation of ecological and evolutionary
timescales (\citeproc{ref-Lion2023}{Lion et al. 2023}) and others
advocating for a more narrow interpretation in which no such separation
is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
al. (\citeproc{ref-Bassar2021}{2021}) identify two types of
eco-evolutionary models that follow from these interpretations. The
first type uses separate equations to model population size change
versus evolutionary change, thereby allowing for any number of
ecological, evolutionary, or environmental feedbacks (e.g.,
\citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
models population demographics as functions of quantitative traits with
ecological and evolutionary change following from demographic processes
and trait distributions operating on the same timescale (e.g.,
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Simmonds2020}{Simmonds et al. 2020};
\citeproc{ref-Jaggi2024}{Jaggi et al. 2024}). Both model types can be
very general, but like all predictive models, they rely on simplifying
assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions are
often grounded in the Price equation to demonstrate accuracy and logical
consistency when modelling evolutionary change (e.g.,
\citeproc{ref-Coulson2008}{Coulson and Tuljapurkar 2008};
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Rees2016}{Rees and Ellner 2016};
\citeproc{ref-Lion2018}{Lion 2018}). For example, Barfield et al.
(\citeproc{ref-Barfield2011}{2011}) link their model back to Price
(\citeproc{ref-Price1970}{1970}), which they consider to be a
``universal law of evolution'', to place their conclusions concerning
stage structured evolution in the broader context of evolutionary
theory. The role of fundamental equations is therefore important for
unifying theory (\citeproc{ref-Luque2021}{Luque and Baravalle 2021}),
and we believe that a fundamental equation of eco-evolutionary change
has been curiously overlooked.

We present an equation from which the fundamental equations of ecology
and evolutionary biology can be derived. Derivation follows by adding
assumptions that are specific to population ecology or evolution in the
same way that key equations of population genetics or quantitative
genetics can be derived from the Price equation by restricting the
domain of interest (e.g., to allele frequencies in the case of
population genetics, or to continuous phenotypes in the case of
quantitative genetics, \citeproc{ref-Queller2017}{Queller 2017}). We
propose our equation as a formal definition of eco-evolutionary change.

@@

\begin{quote}
\textbf{Box 1:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\citeproc{ref-Price1970}{Price 1970},
\citeproc{ref-Price1972}{1972}), it postulates some basic properties
that all evolutionary systems must satisfy: change over time, ancestor
and descendant relations, and a character or phenotype
(\citeproc{ref-Rice2004}{Rice 2004}). Using simple algebraic language,
the Price equation represents evolutionary change with the predominant
notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's fitness, and \(\bar{w}\) average population fitness. On the
right-hand side of the equation, the first term is the covariance
between a character value \(z\) and fitness \(w\), which reflects
\(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Okasha2006}{Okasha 2006}; \citeproc{ref-Frank2012}{Frank
2012}). A more specific version of the covariance term was already known
within the quantitative and population genetics tradition
(\citeproc{ref-Robertson1966}{Robertson 1966}), usually representing the
action of natural selection. The Price equation adds an expectation term
and abstracts away from any specific mechanisms of replication or
reproduction, or mechanisms of inheritance. Its definitional nature and
lack of substantive biological assumptions has been portrayed both as a
strength (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Frank2012}{Frank 2012}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Baravalle2022}{Baravalle and Luque 2022}), and its
greatest weakness. The abstract nature of the Price equation places it
at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fisher's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
\citeproc{ref-Lehtonen2018}{2018}, \citeproc{ref-Lehtonen2020a}{2020};
\citeproc{ref-Queller2017}{Queller 2017}). This abstractness is also key
to developing a more general view of evolution
(\citeproc{ref-Rice2020}{Rice 2020}; \citeproc{ref-Luque2021}{Luque and
Baravalle 2021}; \citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In
contrast, some researchers consider the Price equation just a triviality
(even tautological), and useless without further modelling assumptions
(\citeproc{ref-VanVeelen2005}{van Veelen 2005};
\citeproc{ref-VanVeelen2012}{van Veelen et al. 2012}). The debate
remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
\citeproc{ref-Baravalle2024}{Baravalle et al. 2025}).
\end{quote}

@@

\begin{quote}
\textbf{Box 2:} The number of individuals in any closed population
(\(N\)) at any given time (\(t + 1\)) is determined by the number of
individuals at \(t\) (\(N_{t}\)), plus the number of births (\(Births\))
minus the number of deaths (\(Deaths\)),
\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
for any closed population. Despite its simplicity, it is a general
equation for defining population change and a starting point for
understanding population ecology. Turchin
(\citeproc{ref-Turchin2001}{2001}) notes that a consequence of this
fundamental equation is the tendency for populations to grow
exponentially (technically geometrically in the above case where time is
discrete). This inherent underlying tendency towards exponential growth
persists even as the complexities of real populations, such as
structure, stochasticity, or density-dependent effects are added to
population models (\citeproc{ref-Turchin2001}{Turchin 2001}). Given the
assumption that all individuals in the population are identical, a per
capita rate of birth, \(Births = bN_{t}\), and death,
\(Deaths = dN_{t}\), can be defined. Rearranging and defining
\(\lambda = 1 + b - d\) gives, \(N_{t+1} = N_{t}\lambda\). Here
\(\lambda\) is the finite rate of increase
(\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that because
\(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in size of
any closed population equals its existing size times its finite rate of
increase.
\end{quote}

@@

This foundation of eco-evolutionary change is a statistical interaction
between the demographic processes of birth and death
(\(\beta_{i} - \delta_{i} + 1\)) and some characteristic of an
individual (\(z_{i} + \Delta z_{i}\)). The value of \(\Omega\) is a
summation of individual characteristics and takes the same units as
\(z\). From eqn 1, we can derive the most fundamental equations of
population ecology and evolutionary biology through an appropriate
interpretation of \(z\). Under more limited interpretations of \(z\), we
can also interpret \(\Omega\) as a metric of ecosystem function.

@@

To recover the general equation for population ecology (Box 2), we
define \(z_{i}\) as the identity of \(i\) belonging to the population.
In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is
one member of the population. In this restricted case, \(z\) is count,
which takes the unit 1 (note that `individual' is a label, not a unit,
see \citeproc{ref-NewellEtAl2019}{Newell and Tiesinga 2019}). Further,
we assume that individuals do not change species by setting
\(\Delta z_{i} = 0\). In this case,

@@

We can now interpret \(\Omega\) as the population size at \(t+1\),
\(N_{t+1}\). Summing up \(\beta_{i}\), \(\delta_{i}\), and current
individuals tallies up the total number of individuals in the next time
step,

@@

We therefore recover the general equation for population ecology (eqn 2)
and the fundamental property of exponential growth in populations (eqn
3, \citeproc{ref-Turchin2001}{Turchin 2001}).

@@

In this definition, the longevity of the individual matters. By
definition, an individual that survives from \(t\) to \(t + 1\) has a
higher fitness than one that dies, even if both have the same
reproductive output. With this definition of fitness (eqn 4), we
substitute,

@@

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Baravalle2022}
Baravalle, L., and V. J. Luque. 2022.
\href{https://doi.org/10.1387/theoria.21940}{Towards a pricean
foundation for cultural evolutionary theory}. Theoria 37:209--231.
